Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-20 manuscript/main.tex
change Figure-> fig.
hack up abstract

\begin{abstract}
The gastrointestinal tract, its resident microorganisms, and the central
nervous system are connected by biochemical signaling, also known as the
"microbiome-gut-brain-axis." Both the human brain and the gut microbiome
have critical developmental windows in the first years of life,
raising the possibility that their development is co-occurring and
likely co-dependent. Emerging evidence implicates gut microorganisms and
microbiota composition in cognitive outcomes and neurodevelopmental
disorders (e.g., autism and anxiety), but the influence of gut microbial
metabolism on typical neurodevelopment has not been explored in detail.
We investigated the relationship of the microbiome with the neuroanatomy
and cognitive function of 381 healthy children, demonstrating that
differences in gut microbial taxa and gene functions are associated with
overall cognitive function and with differences in the size of
multiple brain regions.
Using a combination of multivariate linear and machine learning (ML) models,
we showed that many species,
including \emph{Alistipes obesi} and \emph{Blautia wexlerae}, were
associated with higher cognitive function, while some
species such as \emph{Ruminococcus gnavus} were more commonly found in
children with low cognitive scores after controlling for sociodemographic factors.
Microbial genes for enzymes involved in the metabolism of
neuroactive compounds, particularly short-chain fatty acids such as
acetate and propionate, were also associated with cognitive function.
In addition, ML models were able to use microbial taxa to predict the
volume of brain regions, and many taxa that were identified
as important in predicting cognitive function also dominated the
feature importance metric for individual brain regions,
and for specific subscales of cognitive function.
For example, \emph{B. wexlerae}
was the most important species in models predicting the size of the parahippocampal region
in both the left and right hemispheres
and was among the top predictors of gross motor and expressive language performance.
Several species from
the phylum \textit{Bacteroidetes}, including GABA-producing
\emph{B. ovatus}, were important for predicting
the size of the left accumbens area, but not the right. These
findings provide potential biomarkers of neurocognition and brain development
and may lead to the future development of targets for early detection and early
intervention.
\end{abstract}

@@

The complex community of microorganisms living in the gastrointestinal tract of infants and children
may affect how their brains develop.

@@

We investigated the co-development of the brain and microbiome in
381 healthy and neurotypically-developing children (172 female) between 40 days and 10 years of age
(Table 1,  Figure 1A, Figure S1)
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), with older children primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-abundance \emph{Prevotella copri} (Figure S2).
Overall subject-to-subject variation in gut microbial genes
and brain volume profiles was similarly driven largely by subject age (Figure 1E-F).

@@

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0099, q \textless{} 0.01),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0115, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0207, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0258, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging (MRI) profiles (R\textsuperscript{2} = 0.258, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from metagenomic sequences, being tightly coupled
(Figure 1H, p \textless{} 0.001). Other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004})
overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (taxa - 20.1\%, p \textless{} 0.01;
gene functions - 15.9\%, p \textless{} 0.05) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models (LMs)
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N\textsubscript{species} = 116 for 0--120 months, 
N\textsubscript{species} = 54 for 0--6 months, N\textsubscript{species} = 136 for 18--120 months).
No species were significantly associated (q \textless{} 0.20) with
cognitive function in children under 6 months old after adjusting for
age and maternal education.
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched
in children with higher cognitive function scores (Figure 2B),
including \emph{Alistipes obesi},
\emph{Asaccharobacter celatus} (also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
\cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B). \textit{Sutterella wadsworthensis} was the only microbial species
that was significantly negatively associated with cognitive function score (q value = 0.165).

Cognitive assessments (MSEL, WPPSI, WISC)
are composites of multiple subscales that assess different cognitive domains.
While the composite scores are broadly analagous accross ages and test modality,
the subscales assessed in each test are different.
To assess whether microbial taxa were associated with
the development of specific cognitive domains,
we fit multivariate linear models for each species present in at least 10\%
of samples for children under 6 months, 18 to 36 months, or all ages
with a concurrent MSEL score (Figure 2C).
Several taxa associated with composite scores (Figure 2B)
were also associated with one or more subscores, including
\textit{F. prausnitzii} and \textit{A. obesi}.
Interestingly, while no taxa were significantly associated
with composite scores in children under 6 months old,
two species from the genus \textit{Streptococcus}
(\textit{S. peroris} and \textit{S. mitis}), as well as
\textit{Bacteroides fragilis}, were positively associated
with expressive language in this age group.
\textit{S. mitis} was also associated with expressive language
when looking at all ages, while \textit{Eisenbergiella massiliensis},
\textit{Anaerotruncus colihominis}, and an unclassified \textit{Blautia}
species were negatively associated. % Add more here if they match brain?

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2D-F) and found that
several metabolic pathways were either significantly enriched or depleted in
children with higher cognitive ability scores.
For example, genes for degrading the 3-carbon SCFA, propionate,
were significantly depleted in children with higher cognitive ability
scores in children under 6m (Table 2; propionate degradation I,
enrichment score (E.S.) = -0.66, corrected p-value (q)
= 0.029), as were genes for synthesizing the 2 carbon SCFA acetate
(acetate synthesis II, E.S. = -0.491, q = 0.059; acetate synthesis
III, E.S. = -0.409, q = 0.133).
Interstingly, in children over 18m,
these metabolic associations were reversed, with both acetate synthesis
and propionate degredation being enriched in children
with higher cognitive scores
(acetate synthesis I, E.S. = 0.192, q = 0.145;
acetate synthesis II, E.S. = 0.248, q = 0.15;
propionate degredation I, E.S. = 0.154, q = 0.178),
as was synthesis of the 4-carbon SCFA butyrate
(butyrate synthesis II, E.S. = 0.414, q = 0.178).
Synthesis of the amino acid
glutamate was also significantly enriched in children with
higher cognitive function scores in this age group (glutamate synthesis II, E.S. = 0.355, 
q = 0.023). Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

@@

FSEA relies on understanding functional relationships between individual
genes. However, because the relationships between individual taxa are
still largely unknown, we turned to Random Forest (RF) models, a family of 
unsupervised non-parametric machine learning (ML) method that enables
the identification of underlying patterns in large numbers of individual features
(here, microbial species).
Previous studies have reported
successful use of RFs for processing highly-dimensional and sparse data
from the domain of genomics
\cite{amaratungaEnrichedRandomForests2008,brieucPracticalIntroductionRandom2018,chenRandomForestsGenomic2012,franzosaGutMicrobiomeStructure2019,stephanRandomForestApproach2015},
along with other works where it was used
to predict cognitive conditions related to Alzheimer's disease in
different scenarios \cite{ardekaniPredictionIncipientAlzheimer2017,velazquezRandomForestModel2021}.
Given that gut microbial profiles, as well as neurocognitive
development, may partially reflect socioeconomic and demographic
factors, we assessed the performance of RF regressors where maternal
education (here used as a proxy of socioeconomic status (SES)),
sex, and age were included as possible predictors, either alone
or in combination with microbial taxonomic profiles (Table 3).

@@

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.13, mean RMSE 13.01),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 18.25).
Species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (Figure
3A-B, Table S1A-B); Both analysis identified 
\textit{F. prausnitzii}, % F. prausz was mentioned before, ~ line 395
\textit{E. eligens}, % E. eligens was mentioned before, ~ line 395
\textit{Parasutterella excrementihominis}, and % Not mentioned before
\textit{A. obesi} as important % A. obesi was mentioned before, ~ lines 120, 390
(q \textless 0.2 for LMs, cumulative importance \textless 60\% for RFs).
Other taxa that were significant in LMs were not identified by RF models
including \textit{Eggerthella} sp. CAG 298,
and \textit{B. intestinihominis}. \textit{E. ramulus}
was just outside our cutoff (64\% cumulative importance). % Flag - 64%?
RF models also identified an additional 48 species
that were important for discriminating cognitive function scores
(Table 3, Figure 3A). Subject age was consistently ranked highly in
feature importance, which could indicate that decision branches based on
microbial taxa have increased purity when considering the subject's age
or that age itself is a useful predictor.

@@

While LMs did not identify any taxa in the 0-6 month range
that were associated with concurrently measured cognitive function,
we reasoned that developmental trajectories that begin
\textit{in utero} may not be immediately altered by
microbial exposures, such that microbial influences early in life
may take time to manifest. To test this hypothesis,
we asked whether RF models trained on taxonomic profiles
in the first year of life
could predict \textit{future} cognitive performance measured (Figure 3B, D; Table 4).
Models were consistently able to learn microbial species that are associated
with cognitive performance at a future visit
(mean time between visits: 21.5 months, S.D. = 9.9 months; Figure 3C,
mean test set correlation: 0.27).

@@

Nearly all important taxa in RF models that predicted cognitive function
in children older than 18 months were also important in
models that predicted future cognitive function, if they were present (Figure 3B, Table S5).
At the same time, many additional species were important for future prediction models
that were not important in the models measuring concurrent cognitive function.
In particular, feature importance for future prediction models
were dominated by two species of \textit{Klebsiella},
\textit{K. pneumoniae} and \textit{K. varicola} (Figure 3D, E).
Many of the other species that ranked highest in feature importance
were also oportunistic pathogens, or previously reported
to be linked with disease (eg. \textit{R. gnavus}, \textit{E. coli}).

@@

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
MSEL (birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (Figure 4A, Table 5), 
namely expressive language ( test-set RMSE 12.05, test-set correlation 0.368), 
gross motor (test-set RMSE 11.07, test-set correlation 0.137) and 
visual reception (test-set RMSE 15.2, test-set correlation 0.333).
Between the subscales, both the predictability (measured by the test-set figures of merit)
and the relative importances of several taxa (Figure 4B) is variable.
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different (Tables S2-S4).
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of expressive language, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{Fusicatenibacter saccharivorans} are
distinctly predictive of gross motor, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception (Supplenmentary Tables S2-4).

@@

Given that individual microbes could be associated with particular domains
of cognitive function and not others, if there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans
(children over 18 months old, N = 124).
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Tables 5, 6, and S6), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.444, Right = 0.441; importance for age, Left = 6.8\%, Right =
7.7\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.267, Right = 0.315; importance for age, Left = 3.6\%, Right = 6.2\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.275; relative age importances = 1.1\%).

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|}
      \hline\hline
      \textbf{Statistic} & \textbf{MAPE} & \textbf{R} \\\hline
      Mean & 0.073 & 0.146 \\
      Standard deviation & 0.029 & 0.131 \\
      Maximum & 0.205 & 0.598 \\
      75th percentile & 0.081 & 0.212 \\
      50th percentile & 0.066 & 0.129 \\
      25th percentile & 0.057 & 0.073 \\
      Minimum & 0.036 & -0.133 \\\hline
    \end{tabular}
    \caption{\label{tab:neuro}  \textbf{RF models on microbial taxa predicting brain regions have generally good performance.}
    Mean absolute proportional error (MAPE) and correlation coeficient (R)
    of test-set preditions are reported accross models for all brain regions.
    Complete list of values can be found on Table S6.}
    \end{center}
\end{table}

@@

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a consistent number of species between 28 and 42 (median = 38, less than 30\% of the input features)
was responsible for half of the cumulative importance,
regardless of model performance (Figure 4C-E, Table S2).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included nine species - 
\emph{Anaerostipes hadrus},
\emph{B. vulgatus},
\emph{B. ovatus},
\emph{F. saccharivorans},
\emph{Ruminococcus torques},
\emph{Eubacterium rectale},
\emph{Coprococcus comes},
\emph{A. celatus} and
\emph{B. wexlerae}, which combined, account for approximately 10\% of the cumulative relative importances,
computed after fitness-weighting on the reported brain segments (Figure 4D-E).

Among these, \emph{A. hadrus} is the variable with 
highest average relative importance along all reported models
(mean = 1.4\%; min = 0.8\%; max = 3.6\%, Figure 4D-E).
\emph{B. wexlerae}, which was important in RF models of cognitive function, 
was also important in models of the entorhinal, fusiform and lingual areas
and had the highest importance in models predicting the size of
both the left and right right parahippocampus.
Another cluster of model importance contained taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains \emph{B. ovatus} and \emph{B. uniformis},
but also \emph{Alistipes finegoldii and S. salivarius}.

>>>>

\begin{abstract}
    Emerging evidence implicates gut microbial metabolism in neurodevelopmental disorders,
    but its influence on typical neurodevelopment has not been explored in detail.
    We investigated the relationship between the microbiome and neuroanatomy and cognition of 381 healthy children,
    demonstrating that differences in microbial taxa and genes are associated
    with overall cognitive function and the size of brain regions.
    Using a combination of statistical and machine learning (ML) models,
    we showed that species including \textit{Alistipes obesi}, \textit{Blautia wexlerae}, and \textit{Ruminococcus gnavus}
    were enriched or depleted in children with higher cognitive function scores.
    Microbial metabolism of short-chain fatty acids was also associated with cognitive function.
    In addition, ML models were able to predict the volume of brain regions from microbial profiles,
    and taxa that were important in predicting cognitive function
    were also important for predicting individual brain regions and specific subscales of cognitive function.
    These findings provide potential biomarkers of neurocognitive development
    and may enable development of targets for early detection and intervention.
\end{abstract}

@@

The complex community of microorganisms living in the guts of infants and children
may affect how their brains develop.

@@

We investigated the co-development of the brain and microbiome in
381 healthy and neurotypically-developing children (172 female) between 40 days and 10 years of age
(table 1, fig. 1A, fig. S1)
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (fig. 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (fig. 1B, E; PCoA axis 1), with older children primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-abundance \emph{Prevotella copri} (fig. S2).
Overall subject-to-subject variation in gut microbial genes
and brain volume profiles was similarly driven largely by subject age (fig. 1C, F).

@@

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (fig. 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(fig. 1G, R\textsuperscript{2} = 0.0099, q \textless{} 0.01),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0115, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (fig. 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0207, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0258, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging (MRI) profiles (R\textsuperscript{2} = 0.258, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from metagenomic sequences, being tightly coupled
(fig. 1H, p \textless{} 0.001). Other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004})
overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (taxa - 20.1\%, p \textless{} 0.01;
gene functions - 15.9\%, p \textless{} 0.05) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models (LMs)
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (fig. 2A, N\textsubscript{species} = 116 for 0--120 months, 
N\textsubscript{species} = 54 for 0--6 months, N\textsubscript{species} = 136 for 18--120 months).
No species were significantly associated (q \textless{} 0.20) with
cognitive function in children under 6 months old after adjusting for
age and maternal education.
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched
in children with higher cognitive function scores (fig. 2B),
including \emph{Alistipes obesi},
\emph{Asaccharobacter celatus} (also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
\cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(fig. 2B). \textit{Sutterella wadsworthensis} was the only microbial species
that was significantly negatively associated with cognitive function score (q value = 0.165).

Cognitive assessments (MSEL, WPPSI, WISC)
are composites of multiple subscales that assess different cognitive domains.
While the composite scores are broadly analagous accross ages and test modality,
the subscales assessed in each test are different.
To assess whether microbial taxa were associated with
the development of specific cognitive domains,
we fit multivariate linear models for each species present in at least 10\%
of samples for children under 6 months, 18 to 36 months, or all ages
with a concurrent MSEL score (fig. 2C).
Several taxa associated with composite scores (fig. 2B)
were also associated with one or more subscores, including
\textit{F. prausnitzii} and \textit{A. obesi}.
Interestingly, while no taxa were significantly associated
with composite scores in children under 6 months old,
two species from the genus \textit{Streptococcus}
(\textit{S. peroris} and \textit{S. mitis}), as well as
\textit{Bacteroides fragilis}, were positively associated
with expressive language in this age group.
\textit{S. mitis} was also associated with expressive language
when looking at all ages, while \textit{Eisenbergiella massiliensis},
\textit{Anaerotruncus colihominis}, and an unclassified \textit{Blautia}
species were negatively associated. % Add more here if they match brain?

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (table 2, fig. 2D-F) and found that
several metabolic pathways were either significantly enriched or depleted in
children with higher cognitive ability scores.
For example, genes for degrading the 3-carbon SCFA, propionate,
were significantly depleted in children with higher cognitive ability
scores in children under 6m (table 2; propionate degradation I,
enrichment score (E.S.) = -0.66, corrected p-value (q)
= 0.029), as were genes for synthesizing the 2 carbon SCFA acetate
(acetate synthesis II, E.S. = -0.491, q = 0.059; acetate synthesis
III, E.S. = -0.409, q = 0.133).
Interstingly, in children over 18m,
these metabolic associations were reversed, with both acetate synthesis
and propionate degredation being enriched in children
with higher cognitive scores
(acetate synthesis I, E.S. = 0.192, q = 0.145;
acetate synthesis II, E.S. = 0.248, q = 0.15;
propionate degredation I, E.S. = 0.154, q = 0.178),
as was synthesis of the 4-carbon SCFA butyrate
(butyrate synthesis II, E.S. = 0.414, q = 0.178).
Synthesis of the amino acid
glutamate was also significantly enriched in children with
higher cognitive function scores in this age group (glutamate synthesis II, E.S. = 0.355, 
q = 0.023). Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

@@

FSEA relies on understanding functional relationships between individual
genes. However, because the relationships between individual taxa are
still largely unknown, we turned to Random Forest (RF) models, a family of 
unsupervised non-parametric machine learning (ML) method that enables
the identification of underlying patterns in large numbers of individual features
(here, microbial species).
Previous studies have reported
successful use of RFs for processing highly-dimensional and sparse data
from the domain of genomics
\cite{amaratungaEnrichedRandomForests2008,brieucPracticalIntroductionRandom2018,chenRandomForestsGenomic2012,franzosaGutMicrobiomeStructure2019,stephanRandomForestApproach2015},
along with other works where it was used
to predict cognitive conditions related to Alzheimer's disease in
different scenarios \cite{ardekaniPredictionIncipientAlzheimer2017,velazquezRandomForestModel2021}.
Given that gut microbial profiles, as well as neurocognitive
development, may partially reflect socioeconomic and demographic
factors, we assessed the performance of RF regressors where maternal
education (here used as a proxy of socioeconomic status (SES)),
sex, and age were included as possible predictors, either alone
or in combination with microbial taxonomic profiles (table 3).

@@

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.13, mean RMSE 13.01),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 18.25).
Species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (fig.
3A-B, table S1A-B); Both analysis identified 
\textit{F. prausnitzii}, % F. prausz was mentioned before, ~ line 395
\textit{E. eligens}, % E. eligens was mentioned before, ~ line 395
\textit{Parasutterella excrementihominis}, and % Not mentioned before
\textit{A. obesi} as important % A. obesi was mentioned before, ~ lines 120, 390
(q \textless 0.2 for LMs, cumulative importance \textless 60\% for RFs).
Other taxa that were significant in LMs were not identified by RF models
including \textit{Eggerthella} sp. CAG 298,
and \textit{B. intestinihominis}. \textit{E. ramulus}
was just outside our cutoff (64\% cumulative importance). % Flag - 64%?
RF models also identified an additional 48 species
that were important for discriminating cognitive function scores
(table 3, fig. 3A-B, E). Subject age was consistently ranked highly in
feature importance, which could indicate that decision branches based on
microbial taxa have increased purity when considering the subject's age
or that age itself is a useful predictor.

@@

While LMs did not identify any taxa in the 0-6 month range
that were associated with concurrently measured cognitive function,
we reasoned that developmental trajectories that begin
\textit{in utero} may not be immediately altered by
microbial exposures, such that microbial influences early in life
may take time to manifest. To test this hypothesis,
we asked whether RF models trained on taxonomic profiles
in the first year of life
could predict \textit{future} cognitive performance measured at least 6 months
after the stool sample was collected (fig. 3C-D, F; table 4).
Models were consistently able to learn microbial species that are associated
with cognitive performance at a future visit
(mean time between visits: 21.5 months, S.D. = 9.9 months; fig. 3G,
mean test set correlation: 0.27).

@@

Nearly all important taxa in RF models that predicted cognitive function
in children older than 18 months were also important in
models that predicted future cognitive function, if they were present (fig. 3B-C, table S5).
At the same time, many additional species were important for future prediction models
that were not important in the models measuring concurrent cognitive function.
In particular, feature importance for future prediction models
were dominated by two species of \textit{Klebsiella},
\textit{K. pneumoniae} and \textit{K. varicola} (fig. 3C, F).
Many of the other species that ranked highest in feature importance
were also oportunistic pathogens, or previously reported
to be linked with disease (eg. \textit{R. gnavus}, \textit{E. coli}).

@@

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
MSEL (birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (fig. 4A, table 5), 
namely expressive language ( test-set RMSE 12.05, test-set correlation 0.368), 
gross motor (test-set RMSE 11.07, test-set correlation 0.137) and 
visual reception (test-set RMSE 15.2, test-set correlation 0.333).
Between the subscales, both the predictability (measured by the test-set figures of merit)
and the relative importances of several taxa (fig. 4B) is variable.
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different (tables S2-S4).
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of expressive language, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{Fusicatenibacter saccharivorans} are
distinctly predictive of gross motor, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception (tables S2-4).

@@

Given that individual microbes could be associated with particular domains
of cognitive function and not others, if there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans
(children over 18 months old, N = 124).
Some brain regions were more readily
predicted by RF models trained on microbial taxa (tables 5, 6, and S6), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.444, Right = 0.441; importance for age, Left = 6.8\%, Right =
7.7\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.267, Right = 0.315; importance for age, Left = 3.6\%, Right = 6.2\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.275; relative age importances = 1.1\%).

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|}
      \hline\hline
      \textbf{Statistic} & \textbf{MAPE} & \textbf{R} \\\hline
      Mean & 0.073 & 0.146 \\
      Standard deviation & 0.029 & 0.131 \\
      Maximum & 0.205 & 0.598 \\
      75th percentile & 0.081 & 0.212 \\
      50th percentile & 0.066 & 0.129 \\
      25th percentile & 0.057 & 0.073 \\
      Minimum & 0.036 & -0.133 \\\hline
    \end{tabular}
    \caption{\label{tab:neuro}  \textbf{RF models on microbial taxa predicting brain regions have generally good performance.}
    Mean absolute proportional error (MAPE) and correlation coeficient (R)
    of test-set preditions are reported accross models for all brain regions.
    Complete list of values can be found on table S6.}
    \end{center}
\end{table}

@@

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a consistent number of species between 28 and 42 (median = 38, less than 30\% of the input features)
was responsible for half of the cumulative importance,
regardless of model performance (fig. 4C-E, table S2).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included nine species - 
\emph{Anaerostipes hadrus},
\emph{B. vulgatus},
\emph{B. ovatus},
\emph{F. saccharivorans},
\emph{Ruminococcus torques},
\emph{Eubacterium rectale},
\emph{Coprococcus comes},
\emph{A. celatus} and
\emph{B. wexlerae}, which combined, account for approximately 10\% of the cumulative relative importances,
computed after fitness-weighting on the reported brain segments (fig. 4D-E).

Among these, \emph{A. hadrus} is the variable with 
highest average relative importance along all reported models
(mean = 1.4\%; min = 0.8\%; max = 3.6\%, fig. 4D-E).
\emph{B. wexlerae}, which was important in RF models of cognitive function, 
was also important in models of the entorhinal, fusiform and lingual areas
and had the highest importance in models predicting the size of
both the left and right right parahippocampus.
Another cluster of model importance contained taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains \emph{B. ovatus} and \emph{B. uniformis},
but also \emph{Alistipes finegoldii and S. salivarius}.
